---
title: "Applied Exercise 10.9"
author: "Team 21: Derartu Dinku | Kelly Field | Sy Lax | Kishan Patel"
output: html_document
---

Consider the USArrests data. We will now perform hierarchical clustering on the states.

a. Using hierarchical clustering with complete linkage and Euclidean distance, cluster the states.

```{r}
set.seed(2)
hc.complete <- hclust(dist(USArrests), method = "complete")
plot(hc.complete)
```

b. Cut the dendrogram at a height that results in three distinct clusters. Which states belong to which clusters?
```{r}
cutree(hc.complete, 3)
```

c. Hierarchically cluster the states using complete linkage and Eu- clidean distance, after scaling the variables to have standard deviation one.
```{r}
sd.data <- scale(USArrests)
hc.complete.sd <- hclust(dist(sd.data), method = "complete")
plot(hc.complete.sd)
```

d. What effect does scaling the variables have on the hierarchical clustering obtained? In your opinion, should the variables be scaled before the inter-observation dissimilarities are computed? Provide a justification for your answer.

```{r}
cutree(hc.complete.sd, 3)
```

```{r}
table(cutree(hc.complete, 3), cutree(hc.complete.sd, 3), dnn = c('Not-Scaled', 'Scaled'))
```

Scaling the variables affect the clusters obtained although by scaling the data the size of the third cluster increases subastantially making up alomost 1/3 of the data present. Varaibles used in clustering algorithms should always be scaled beforehand. The diffence in scale of the features used  could artificially pull a data point towards a particular cluster.